
\documentclass[copyright,creativecommons]{eptcs} 
\providecommand{\event}{ICE 26} % Name of the event you are submitting to
\usepackage{underscore}           % Only needed if you use pdflatex.
 
\usepackage{breakurl}

\usepackage{wrapfig}

\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsbsy}
\usepackage{enumerate}
\usepackage{stackrel}
\usepackage{bm}
%\usepackage[toc,page]{appendix}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{color}
\usepackage[utf8]{inputenc}
\usepackage{dashbox}
%\usepackage{pgfkeys}
\usepackage[capitalise]{cleveref}
\usepackage{accents}
\crefformat{enumi}{condition~#2#1#3}
\crefname{fact}{Fact}{Facts}
\Crefname{fact}{Fact}{Facts}
\crefformat{fact}{Fact~#2#1#3}

%%% The following setting
%%% enable figures not to float too much
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.7}
%%%%%%%%%%%%%%%%%%%
%%%%%
%%%%%%%%
%
%
%
%\usepackage{derivation}
%
%
\usepackage{mathtools}
%\usepackage{stmaryrd}
%
%\usepackage{pifont}
%\newcommand{\cmark}{\mbox{\tiny \ding{51}}}%
%\newcommand{\xmark}{\mbox{\tiny \ding{55}}}%
%%\newcommand{\cmark}{\mbox{\tt ok}}% 
%%\newcommand{\xmark}{\mbox{\tt no}}%
%
%\newcommand{\Der}{{\cal D}}
%
%
%
%
%\bibliographystyle{plainurl}
%
%%\usepackage{cmll}
%
\input{ggstyle}
%
\input{macros}

%\modulolinenumbers[5]
\def \bmr {\begin{color}{red}} 
\def \emr {\end{color}}
\def \bfr {\begin{color}{Fuchsia}} 
\def \efr {\end{color}}
\def \bmc {\begin{color}{magenta}Mariangiola: } 
\def \emc {\end{color}}
\def \bfc {\begin{color}{brown}Franco: } 
\def \efc {\end{color}}
\def \brc {\begin{color}{blue}Rolf: } 
\def \erc {\end{color}}
\def \brr {\begin{color}{olive}} 
\def \err {\end{color}}

 
\author{
%Franco Barbanera\thanks{ Partially supported by 
%Project “National Center for HPC, Big Data e Quantum Computing”,  Programma M4C2, Investimento 1.3.}
%\institute{Dipartimento di Matematica e Informatica\\
%University of Catania}
%\email{franco.barbanera@unict.it}
%\and 
%Rolf Hennicker
%\institute{Institute for Informatics\\
%LMU Munich}
%\email{hennicke@pst.ifi.lmu.de}
}




\begin{document}

%\title{Multicomposition
% of Systems \\ of Communicating Finite State Machines
%}

\title{Title%Partial Fusion
}

\def\titlerunning{Partial Fusion}
\def\authorrunning{
%F.\,Barbanera \& R.\,Hennicker
}

\maketitle

\begin{abstract}
The {\em Participants-as-Interfaces\/} (PaI) approach to system composition 
suggests that participants of a system may be viewed as interfaces.
Given a set of systems,
% of communicating finite state machines (CFSMs),  
one participant per system is chosen to play the role of an interface. When systems are composed, the %each CFSM
% of an 
interface participants are replaced by  \emph{gateways} which communicate to each other
by forwarding messages.
We generalise the approach to 
\emph{partial gateways}, where 
gateways can forward only a choosen
set of messages. 
As for the standard PaI approach, we exploit such extended version for
systems of asynchronous  communicating finite state machines (CFSMs).
We prove that a bunch of relevant 
communication properties (deadlock-freeness, reception-error-freeness, etc.) are preserved by {\em PaI composition via partial gateways\/} in case also the
connection policy (i.e. the system representing the way we wish gateways interact with each other) enjoys the same properties.
Such a proof turns out to be just a corollary of a preservation result for a restricted
and binary version of PaI composition
via partial gateways, dubbed \emph{fusion-composition}.
In fact any 
PaI composition (binary, with multiple systems, orchestrated) simply reduces
to a number of binary fusion compositions.
Fusion-composition hence turns out to be at the heart of the PaI approach.
\end{abstract}


\input{Introduction}
\input{cfsm}
\input{comp-via-partial-gw}
\input{safeness-of-binary-partial-fusion}
\input{preservation-results}

%\paragraph*{\bf Acknowledgements}
% We warmly thank the ICE'24 reviewers for their careful reading, their thoughtful comments/suggestions and the helpful discussion in the forum. We also thank Emilio Tuosto for his nice tikz style for automata.
%
%
\setlength{\abovedisplayskip}{6pt}
\setlength{\belowdisplayskip}{\abovedisplayskip}

\bibliographystyle{eptcs}
\bibliography{session}

%------APPENDIX--------
%\newpage
%\appendix
%
%\input{safenessPreservation}
%\input{weakDeadlockPreservation}
%\input{strong-deadlock-free-Preservation}
%---------------------------













%------------------------------------------------------------------------------
% Index
%\printindex

%------------------------------------------------------------------------------
\end{document}

% EOF
